WORCS: A Workflow for Open Reproducible Code in Science

### 2020-11-03 paper/Manuscript_DS.Rmd
R1 C9 Should everyone use R?

  affiliation: | 
      a: Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics  
      b: Open Science Community Utrecht  
      c: Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany  
      d: Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK  
      e: UMC Utrecht  
      f: Utrecht University, Department of Information and Computing Sciences  
      g: Humboldt-Universitaet zu Berlin  
      h: Utrecht University faculty of Humanities, department of Language, Literature & Communication, Utrecht Institute of Linguistics OTS  
      i: University Library Utrecht
title: "WORCS: A Workflow for Open Reproducible Code in Science"
short:      "WORKFLOW FOR OPEN SCIENCE"
shorttitle:      "WORKFLOW FOR OPEN SCIENCE"
abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible, accompanied by an open source software implementation that 
  automates these steps.
  The 
  purpose of the workflow is to lower the threshold for 
  adoption of open science principles. 
  It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  This workflow is primarily targeted towards
  scholars working in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and (optionally) tabular data - but it is flexible enough to accommodate many
  other scenarios.
  To facilitate widespread adoption, the
  WORCS principles are implemented in the R package `worcs`, which offers an
  RStudio 
  project template that is, effectively, a "one-click solution" for R users,
  and utility functions for specific workflow steps.
  This paper addresses the design of the workflow,
  discusses how it meets different 
  standards for open science, and provides examples of the implementation of
  `worcs` in `R`. The source code for the `R` package and manuscript,
  and a list of user examples of WORCS projects, are available at 
  https://github.com/cjvanlissa/worcs.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rmarkdown::html_document
---
<!--
, "worcs_packages.bib"
output: 
  rmarkdown::html_document:
    citation_package: biblatex Note: Citations are always handled by pandoc when format is not PDF
knit: worcs::cite_all
csl: apa.csl
csl: apa.csl
-->

@@

Many researchers are motivated to adopt current best practices for open
science and enjoy these benefits. And yet, the question of how to 
get started can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by presenting a simple workflow 
that meets most requirements open science, and is based on best practices:
The *Workflow for Open Reproducible Code in Science* (WORCS).
WORCS is a lightweight approach to open science and
computational reproducibility for scholars working in `R` on a private computer,
conducting research that involves academic prose, analysis code,
and (optionally) tabular data - but it is flexible enough to accommodate many
other scenarios. The workflow (Figure 1) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project open and reproducible.
WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them.
It can also be used to produce work in accordance with best practices
in the absence of top-down support or guidelines.
Although the principles underlying this workflow are universal, WORCS has been
implemented in `R` [@rcoreteamLanguageEnvironmentStatistical2020].
The `worcs` package [@worcspackage2020] offers a project template for RStudio
[@rstudioteamRStudioIntegratedDevelopment2015], which is, effectively,
a "one click solution". This paper discusses the principles underlying WORCS,
introduces the workflow, addresses how WORCS meets best practices for open science,
and illustrates the use of the `worcs` package.

@@

WORCS is, first and foremost, a conceptual workflow that could be implemented in any software environment.
As of this writing, the workflow has been implemented for `R` users in the package `worcs` [@worcspackage2020].
Several arguments support the choice to implement this workflow in `R` and RStudio first, although we encourage developers to port the software to other languages.
First, `R` and all of it extensions are free open source software, which
make it a tool of choice for open science.
Second, all tools required for an open science workflow are implemented in `R`,
most of these tools are directly accessible through the RStudio user interface,
and some of them are actively developed by the team behind RStudio.
Third, RStudio is a Public Benefit Corporation, whose commitment to open science is codified in the company charter:

@@

The second solution is to share a synthetic dataset with similar characteristics
to the real data. Synthetic data mimic the level of measurement and
(conditional) distributions of the real data
[see @nowokSynthpopBespokeCreation2016]. Sharing synthetic data allows any third
party to 1) verify that the published code works, 2) debug the code, and 3)
write valid code for alternative analyses. It is important to note that complex
multivariate relationships present in the real data are often lost in synthetic
data. Thus, findings from the real data might not be replicated in the synthetic
data, and findings in the synthetic data *should not be substantively interpreted*.
Still, sharing synthetic data facilitates data requests from third parties. A
third party can write analysis code based on the synthetic data, and send it to
the authors who evaluate it on the real data and send back the results. The
`worcs` package offers a simple but flexible function to generate synthetic data,
`synthetic()`, which by default uses random forests to generate a model for
synthesis (see @wrightRangerFastImplementation2015). For data synthesis to work
properly, it is essential that the `class` (data type) of variables is defined
correctly. As of `worcs` version `r worcs:::cran_version()`, numeric, integer, factor, and logical data
are supported out of the box. Other types of variables should be converted to
one of these types, or users can use a custom `model_expression` when calling
`synthetic`.

@@

In `worcs`, researchers can call the function `open_data(data)` to make the data
publicly available. This function stores the `data` object (such as a `data.frame` or
`matrix`) in a tabular data file (called `data.csv` by default).
It also generates an
accompanying codebook - 
a human-readable document that serves as a legend for the tabular data file -
that can be viewed online.
All changed files should now be added to the Git repository, committed, and
pushed to the remote repository.
This can be done through the Git panel in RStudio, or by running the `worcs`
function `git_update("commit message")`.
Once the remote repository is made public, these data will be
open to the public.
Assume that, once this is done, the data cannot be un-shared.

Alternatively, if the project requires data to remain closed, researchers can
call `closed_data(data)`. This function also stores the data in a local `.csv` file
and generates a codebook,
but the original data file is added to `.gitignore` so it cannot be
accidentally added to the Git repository.
The function also computes a checksum for the original data, and logs it in the
`.worcs` project file. This checksum serves as the indelible record of the state 
of the raw data.
The function `closed_data()` also creates a synthetic dataset (called
`synthetic_data.csv` by default).
The user should push all changed files to the remote repository.
Once the remote repository is made public, people will have access to a
checksum for the original data that exist on your local machine, a codebook
describing those data, and a synthetic copy of the data.
Keep in mind that, when using closed data, the original data file exists only on
the user's computer. Adequate provisions to back up the data, while respecting
principles of responsible data stewardship, should be made.

>>>>

  affiliation: | 
      a: Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics  
      b: Open Science Community Utrecht  
      c: Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany  
      d: Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK  
      e: UMC Utrecht  
      f: Utrecht University, Department of Information and Computing Sciences  
      g: Humboldt-Universitaet zu Berlin  
      h: Utrecht University faculty of Humanities, department of Language, Literature & Communication, Utrecht Institute of Linguistics OTS  
      i: University Library Utrecht
title: "WORCS: A Workflow for Open Reproducible Code in Science"
short:      "WORKFLOW FOR OPEN SCIENCE"
shorttitle:      "WORKFLOW FOR OPEN SCIENCE"
abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible.
  The purpose of the workflow is to lower the threshold for 
  adoption of open science principles.
  It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  To facilitate widespread adoption, the WORCS principles have been implemented
  in the R package `worcs`, which offers an RStudio 
  project template and utility functions for specific workflow steps.
  This paper introduces the conceptual workflow,
  discusses how it meets different 
  standards for open science, and provides examples of the implementation in `R`. 
  This paper is primarily targeted towards
  scholars conducting research projects in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and tabular data that does not require specialized computational
  provisions.
  However, the workflow is flexible enough to accommodate other scenarios,
  and offers a sensible starting point for customized solutions.
  The source code for the `R` package and manuscript,
  and a list of user examples of WORCS projects, are available at 
  https://github.com/cjvanlissa/worcs.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rmarkdown::html_document
---
<!--
, "worcs_packages.bib"
output: 
  rmarkdown::html_document:
    citation_package: biblatex Note: Citations are always handled by pandoc when format is not PDF
knit: worcs::cite_all
csl: apa.csl
csl: apa.csl
-->

@@

Many researchers are motivated to adopt current best practices for open
science and enjoy these benefits. And yet, the question of how to 
get started can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by presenting a simple workflow,
based on best practices,  
that meets most requirements open science:
The *Workflow for Open Reproducible Code in Science* (WORCS).

This paper introduces the conceptual workflow (referred to in upper case, "WORCS"),
and discusses its underlying principles and how it meets different standards for open science.
The principles underlying this workflow are universal and largely platform independent.
We also present a software implementation of WORCS for `R` users [@rcoreteamLanguageEnvironmentStatistical2020]:
The `R` package `worcs` [referred to in monospace font, @worcspackage2020].
This package offers a project template for RStudio
[@rstudioteamRStudioIntegratedDevelopment2015],
and several convenience functions to automate specific workflow steps.
This paper is most relevant for scholars using R on a private computer to
conducting research that involves academic prose, analysis code,
and tabular data.
For other readers, it can serve as a primer on open science workflows,
and provide a sensible starting point for a customized solution.

The WORCS workflow (Figure 1) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project open and reproducible.
WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them.
It can also be used by individual authors to produce work in accordance with best practices
in the absence of top-down support or guidelines (i.e., "grass roots" open science).

@@

WORCS is, first and foremost, a conceptual workflow that could be implemented in any software environment.
As of this writing, the workflow has been implemented for `R` users in the package `worcs` [@worcspackage2020].
Several arguments support our choice to implement this workflow first in `R` and RStudio,
although we encourage developers to port the software to other languages.
First, `R` and all of it extensions are free open source software, which
make it a tool of choice for open science.
Second, all tools required for an open science workflow are implemented in `R`,
most of these tools are directly accessible through the RStudio user interface,
and some of them are actively developed by the team behind RStudio.
Third, RStudio is a Public Benefit Corporation, whose commitment to open science is codified in the company charter:

@@

The second solution is to share a synthetic dataset with similar characteristics
to the real data. Synthetic data mimic the level of measurement and
(conditional) distributions of the real data
[see @nowokSynthpopBespokeCreation2016]. Sharing synthetic data allows any third
party to 1) verify that the published code works, 2) debug the code, and 3)
write valid code for alternative analyses. It is important to note that complex
multivariate relationships present in the real data are often lost in synthetic
data. Thus, findings from the real data might not be replicated in the synthetic
data, and findings in the synthetic data *should not be substantively interpreted*.
Still, sharing synthetic data facilitates data requests from third parties. A
third party can write analysis code based on the synthetic data, and send it to
the authors who evaluate it on the real data and send back the results. The
`worcs` package offers a simple but flexible function to generate synthetic data,
`synthetic()`, which by default uses random forests to generate a model for
synthesis (see @wrightRangerFastImplementation2015).

Note that for data synthesis to work
properly, it is essential that the `class` (data type) of variables is defined
correctly. As of `worcs` version `r worcs:::cran_version()`, numeric, integer, factor, and logical data
are supported out of the box. 
Other types of variables should be converted to one of these types.
It is important to consider this when using special data types,
such as the labeled vectors created by the `haven` package when reading an SPSS file.
We further caution that the default random forests algorithm is computationally intensive,
and may be unfeasible for use with "large" datasets.
Users can provide use a custom `model_expression` and `predict_expression`
to use a different algorithm when calling `synthetic()`.
For example, users could use `lm()` as a `model_expression` to use linear regression,
which preserves linear marginal relationships but can give rise to values out of range of the original data.
Or users could call `sample()` as a `predict_expression` to bootstrap each variable,
a very quick solution that maintains univariate distributions but loses all marginal relationships.

@@

In `worcs`, researchers can call the function `open_data(data)` to make the data
publicly available. This function stores the `data` object (such as a `data.frame` or
`matrix`) in a tabular data file.
It also generates an
accompanying codebook - 
a human-readable document that serves as a legend for the tabular data file -
that can be viewed online.
All changed files should now be added to the Git repository, committed, and
pushed to the remote repository.
This can be done through the Git panel in RStudio, or by running the `worcs`
function `git_update("commit message")`.
Once the remote repository is made public, these data will be
open to the public.
Assume that, once this is done, the data cannot be un-shared.

Alternatively, if the project requires data to remain closed, researchers can
call `closed_data(data)`. This function also stores the data in a local `.csv` file
and generates a codebook,
but the original data file is added to `.gitignore` so it cannot be
accidentally added to the Git repository.
The function also computes a checksum for the original data, and logs it in the
`.worcs` project file. This checksum serves as the indelible record of the state 
of the raw data.
The function `closed_data()` also calls the aforementioned `synthetic()` function to
create a synthetic dataset.
The user should push all changed files to the remote repository.
Once the remote repository is made public, people will have access to a
checksum for the original data that exist on your local machine, a codebook
describing those data, and a synthetic copy of the data.
Keep in mind that, when using closed data, the original data file exists only on
the user's computer. Adequate provisions to back up the data, while respecting
principles of responsible data stewardship, should be made.

### 2020-11-03 paper/reviews_ds.Rmd
R1 C9 Should everyone use R?

\RC{In some cases it looks like the authors claim that everyone should use R for open science. While R has a strong community behind, I believe there are others with very strong support (e.g. Python). I am not sure if suggesting a shift to R is the right approach. Note that this is the impression I got from reading the paper, maybe it was unintended by the authors.}

Response: Clearly define scope; feedback is welcome.

\Assignment{All}
\WorkInProgress
\Hard

>>>>

\RC{C9: In some cases it looks like the authors claim that everyone should use R for open science. While R has a strong community behind, I believe there are others with very strong support (e.g. Python). I am not sure if suggesting a shift to R is the right approach. Note that this is the impression I got from reading the paper, maybe it was unintended by the authors.}

Response: We wish to emphasize that it was certainly unintended to claim that everyone should us R for open science. To address this comment, we have done two things: First, throughout the paper, we have attempted to clarify the distinction between the platform independent conceptual workflow "WORCS", and its implementation for R users, `worcs`. Second, throughout the paper, we encourage the implementation of WORCS in other languages. Most illustrative of these changes is the start of the paragraph on **The R implementation of WORCS**:

> WORCS is, first and foremost, a conceptual workflow that could be implemented in any software environment. As of this writing, the workflow has been implemented for `R` users in the package `worcs` [@worcspackage2020]. Several arguments support our choice to implement this workflow first in `R` and RStudio, although we encourage developers to port the software to other languages.
